---
title: "Play style evolution of the top 4 teams in Europe"
author: "JT"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
---

A vibecoding project on visualising how the playstyle of the top four teams in the major European leagues (that compete in the Champions league) have evolved since 2017.

Let us first install and load the libraries needed.
```{r packages} 
# install.packages(c("tidyverse", "gganimate", "gifski"))

library(tidyverse)
library(gganimate)
library(gifski)
```

# 1. Loading the data

I have already saved the data from the previous exercise of using python and the premier_league library, so let us load that up. 

```{r data}
file_path <- "data/data.csv"
df <- read.csv(file_path)
```

# 2. The dataset

Instead of simply filtering to top 4, we keep ALL season data for any team that was ever in the top 4. We then flag each row as top 4 or not, and map that to alpha. This way gganimate can interpolate positions for teams entering or exiting the top 4 — they slide smoothly while fading in/out.

```{r filter}
# Identify every team that finished top 4 in at least one season
ever_top4 <- df %>%
    filter(position <= 4) %>%
    distinct(team) %>%
    pull(team)

# Keep all seasons for those teams, flag top-4 status per row
df_animated <- df %>%
    filter(team %in% ever_top4) %>%
    mutate(
        in_top4     = position <= 4,
        point_alpha = if_else(in_top4, 0.85, 0),
        label_alpha = if_else(in_top4, 1.0,  0)
    )

# Reference subset for axis limits (keeps the view focused on top-4 range)
df_top4 <- df_animated %>% filter(in_top4)
```

# 3. The base ggplot chart

```{r base chart}
animated_plot <- ggplot(df_animated, aes(x = avg_progressive_passes,
                                         y = avg_possession_rate,
                                         color = league,
                                         group = team)) +

    # Scatterplot points — alpha driven by top-4 status
    geom_point(aes(alpha = point_alpha), size = 3) +

    # Team text labels — also fade with dots
    geom_text(aes(label = team, alpha = label_alpha),
              nudge_y = 1.8,
              size = 2.2,
              family = "sans",
              check_overlap = FALSE,
              show.legend = FALSE) +

    # Treat alpha values as literal (0 = invisible, 0.85 = visible)
    scale_alpha_identity() +

    # Top Right Label ("Possession-based")
    annotate("text", x = Inf, y = Inf, label = "Possession-based",
             hjust = 1.05, vjust = 1.5, size = 6,
             family = "sans", fontface = "bold", color = "grey70", alpha = 0.8) +

    # Bottom Left Label ("Counter-attacking")
    annotate("text", x = -Inf, y = -Inf, label = "Counter-attacking",
             hjust = -0.05, vjust = -0.5, size = 6,
             family = "sans", fontface = "bold", color = "grey70", alpha = 0.8) +

    scale_color_brewer(palette = "Set1") +
    theme_minimal(base_size = 14) +
    theme(
        text = element_text(family = "sans"),
        legend.position = "right",
        plot.title = element_text(face = "bold", size = 18),
        plot.subtitle = element_text(color = "grey30", size = 15, face = "italic"),
        panel.grid.minor = element_blank()
    ) +

    # Axis limits anchored to the top-4 data range
    coord_cartesian(
        xlim = c(min(df_top4$avg_progressive_passes, na.rm=TRUE) - 2,
                 max(df_top4$avg_progressive_passes, na.rm=TRUE) + 2),
        ylim = c(min(df_top4$avg_possession_rate, na.rm=TRUE) - 2,
                 max(df_top4$avg_possession_rate, na.rm=TRUE) + 4)
    ) +

    labs(
        title = 'Tactical Evolution: Top 4 Teams across Europe',
        subtitle = 'Season: {closest_state}',
        x = 'Average Progressive Passes per Match',
        y = 'Average Possession Rate (%)',
        color = 'League',
        caption = "Source: premier_league"
    ) +

    # Motion logic — gganimate now interpolates both position AND alpha
    transition_states(
        states = season,
        transition_length = 1,
        state_length = 1,
        wrap = FALSE
    ) +

    # Fade for teams that only appear in the very first/last season
    enter_fade() +
    exit_fade()

```

# 4. The animation

```{r animation}
output_gif <- "output/tactical_evolution_top4.gif"

animate(
    animated_plot,
    nframes = 580,
    fps = 20,
    width = 900,
    height = 650,
    units = "px",
    res = 150,
    end_pause = 60,
    renderer = gifski_renderer(output_gif)
)

```

We can see that there is variety in terms of playstyle across the top four teams in each major European league, so there is no **exceptionally dominant** playstyle per se and we should be happy to see that football is ever-evolving! Also interestingly, we can see teams that have made significant playstyle shifts (i.e. Newcastle and Bayer Leverkusen).
